Nucleotide Diversity Loss on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-11-16 manuscript/Ydiversity_nov12.tex
minor edits

\section*{Results and Discussion}
\subsection*{Extensive loss of Y-chromosome diversity}
Our analysis of polymorphism levels across the genome in the dioecious plant \textit{R. hastatulus} reveals that neutral diversity on the Y chromosome is substantially lower than expected from a standard neutral model. In particular, neutral diversity on Y-linked genes $\theta_{Y}$ is on average \textasciitilde 2.1\% of the value for their homologues on the X chromosome. Talking 1/4 of the mean autosomal diversity as the equilibrium expectation under neutrality \citep{wright1931evolution}, this corresponds to a chromosome-wide reduction of \textasciitilde 93\%. Note that by comparing X and Y diversity to autosomal diversity, our results indicate that the difference between X and Y homologues is not due to an elevation of diversity on the X chromosome, but rather a Y-specific reduction. Our results do suggest a slight elevation in X-linked diversity relative to the neutral prediction of X/A=0.75, but this difference is not significantly different from the expectation for the empirically-estimated ratio of 0.6. Although our estimates of diversity have not beed normalized by divergence, previous work has shown that the average synonymous substitution rate, between \textit{R. hastatulus} and the non-dioecious outgroup \textit{R. bucephalophorus} is not significantly different for sex-linked (0.2016) or autosomal genes (0.219) \citep{hough2014}. It is therefore unlikely that our results are caused by mutation rate differences between sex-linked and autosomal genes.

\subsection*{Female biased sex ratios and variance in male offspring number}

>>>>

\section*{Results and Discussion}
\subsection*{Extensive loss of Y-chromosome diversity}
Our analysis of polymorphism levels across the genome in the dioecious plant \textit{R. hastatulus} reveals that neutral diversity on the Y chromosome is substantially lower than expected from a standard neutral model. In particular, neutral diversity on Y-linked genes $\theta_{Y}$ is on average \textasciitilde 2.1\% of the value for their homologues on the X chromosome. Talking 1/4 of the mean autosomal diversity as the equilibrium expectation under neutrality \citep{wright1931evolution}, this corresponds to a chromosome-wide reduction of \textasciitilde 93\%. Note that by comparing X and Y diversity to autosomal diversity, our results indicate that the difference between X and Y homologues is not due to an elevation of diversity on the X chromosome, but rather a Y-specific reduction. Our results do suggest a slight elevation in X-linked diversity relative to the neutral prediction of X/A=0.75, but this difference is not significantly different from the expectation for the empirically-estimated ratio of 0.6. Although our estimates of diversity have not beed normalized by divergence, previous work has shown that the average synonymous substitution rate, between \textit{R. hastatulus} and the non-dioecious outgroup \textit{R. bucephalophorus} is not significantly different for sex-linked (0.2016) or autosomal genes (0.219) \citep{hough2014}. It is therefore unlikely that our results are caused by mutation rate differences between sex-linked and autosomal genes. We next consider several possible models - neutral and selective - that might explain the loss of Y chromosome diversity.

\subsection*{female-biased sex ratios and variance in male offspring number}
The occurrence of female-biased sex ratios in \textit{R. hastatulus} is expected to lower Y diversity through lowering male $N_{e}$ and therefore the $N_{e}$ of the Y chromosome. In addition, the $N_{e}_{Y}$ might be further reduced due to high variance in male reproductive success (Figure 1), which is not unusual in annual wind-pollinated plants such as \textit{R. hastatulus} that exhibit extensive phenotypic plasticity in plant size and flower production (\X Harper 1977). Given that male plants in this species produce large amounts of pollen, and female flowers are uniovulate, there is reason to believe that there could be strong competition among males.
